### 2017-11-15 paper/sections/1_introduction.tex
paper: addresses #20 and #21
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>

\begin{figure}[t]
\noindent\includegraphics[width=0.5\textwidth]{figures/cache-management.png}\\

\caption{Using our data management language and Mantle policy engine, we
designed a dynamically sized caching policy for the ParSplice application. The
solution uses knowledge about the application to detect key access patterns and
adjust the cache accordingly. 
\label{fig:cache-management}}
\end{figure}

>>>>

\begin{figure}[t]
\noindent\includegraphics[width=0.5\textwidth]{figures/cache-management.png}\\
\caption{Using our data management language and policy engine, we designed a
dynamically sized caching policy (dashed line) for ParSplice. The solution uses
knowledge about the application to detect key access patterns and adjust the
cache accordingly. Compared to existing ParSplice configurations (solid lines),
our solution saves the most memory without sacrificing performance.
\label{fig:cache-management}}
\end{figure}

### 2017-11-15 paper/sections/2_parsplice-keyspace-analysis.tex
paper: typos
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>
paper: addresses #20 and #21
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>

\begin{figure}[t]
\noindent\includegraphics[width=0.5\textwidth]{figures/parsplice.png}\\
\caption{The ParSplice architecture has a storage hierarchy of caches and a
dedicated cache node (boxes) backed by a persistent database (DB). A splicer
(S) tells worker (W) to generate segments and workers employ tasks (T) for more
parallelization. We focus on the worker's cache (circled), which facilitates
communication and segment exchange between the worker and its
\label{fig:parsplice}}
\end{figure}

@@

\subsubsection*{Entropy increases over time} The reads per second in
Figure~\ref{fig:motivation-regimes} show that the number of requests decreases
and the number of active keys increases over time. The resulting key access
imbalance for the two growth rates in Figure~\ref{fig:motivation-regimes} are
shown in Figure~\ref{fig:methodology-keys}, where reads are plotted for each
unique state, or key, along the \(x\) axis. Keys are more popular than others
(up to \(5\times\)) because worker tasks start generating states with different
coordinates later in the run (this motivates Section~\S\ref{sec:dom-specific}).
The growth rate, temperature, and number of workers have a predictable effect
on the structure of the keyspace.  Figure~\ref{fig:methodology-keys} shows that
the number of reads changes with different initial conditions (\(\Delta\)), but
that the spatial locality of key accesses is similar ({\it e.g.}, some keys are
still \(5\times\) more popular than others).
Figure~\ref{fig:motivation-regimes} shows how entropy for different growth
rates has temporal locality, as the reads per second for \(\Delta_2\) looks
like the reads per second for \(\Delta_1\) stretched out along the time axis.
Trends also exist for temperature and number of workers but are omitted here
for space. This structure means that we can learn the regimes and adapt the
storage system (this motivates Section~\S\ref{sec:dom-specific}).

>>>>

\begin{figure}[t]
\noindent\includegraphics[width=0.5\textwidth]{figures/parsplice.png}\\
\caption{The ParSplice architecture has a storage hierarchy of caches and a
dedicated cache node (boxes) backed by a persistent database (DB). A splicer
(S) tells worker (W) to generate segments and workers employ tasks (T) for more
parallelization. We focus on the worker's cache (circled), which facilitates
communication and segment exchange between the worker and its tasks.
\label{fig:parsplice}}
\end{figure}

@@

\subsubsection*{Entropy increases over time} The reads per second in
Figure~\ref{fig:motivation-regimes} show that the number of requests decreases
and the number of active keys increases over time. The resulting key access
imbalance for the two growth rates in Figure~\ref{fig:motivation-regimes} are
shown in Figure~\ref{fig:methodology-keys}, where reads are plotted for each
unique state, or key, along the \(x\) axis. Keys are more popular than others
(up to \(5\times\)) because worker tasks start generating states with different
coordinates later in the run (this motivates Section~\S\ref{sec:dom-specific}).
The growth rate, temperature, and number of workers have a predictable effect
on the structure of the keyspace.  Figure~\ref{fig:methodology-keys} shows that
the number of reads changes with different initial conditions (\(\Delta\)), but
that the spatial locality of key accesses is similar ({\it e.g.}, some keys are
still \(5\times\) more popular than others).
Figure~\ref{fig:motivation-regimes} shows how entropy for different growth
rates has temporal locality, as the reads per second for \(\Delta_2\) looks
like the reads per second for \(\Delta_1\) stretched out along the time axis.
Trends also exist for temperature and number of workers but are omitted here
for space. This structure means that we can learn the key access patterns and adapt the
storage system (this motivates Section~\S\ref{sec:dom-specific}).

### 2017-11-15 paper/sections/3_methodology.tex
paper: addresses #14 and #15
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>
paper: addresses #20 and #21
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>

%  \item background: CephFS implementation
To explore dynamic cache management policies that change during the run, we use
the data management language and policy engine presented
in~\cite{sevilla:sc15-mantle}. The prototype in that paper was called Mantle
and it was originally built on the CephFS so administrators could control file
system metadata data management policies. We now refer to Mantle as a policy
engine that supports our data management language.  The basic premise is that
data management policies can be expressed with a simple API consisting of
``when", ``where", and ``how much". ``when" controls how aggressive or
conservative the decisions are; ``where" lets the policy control how
distributed or concentrated the data should be; and ``how much" controls the
amount of data that should be sent.

The succinctness of the API lets users inject multiple, dynamic policies. In
this work we focus on a single node, so the ``where" policy is not used. When
we move ParSplice to a distributed key-value store back-end, the ``where" policy
can be used determine which key-value pairs should be moved to which node.

@@

% env of metrics
Using Mantle cluster metrics, we expose cache size, CPU utilization, and memory
pressure of the worker node to the cache management policies. In
Section~\S\ref{sec:parsplice-keyspace-analysis} we only end up using the cache
size although the other metrics proved to be valuable debugging tools. Using
Mantle time series metrics, we expose accesses to the cache as a list of
\texttt{timestamp, keyid} pairs. In Section~\S\ref{sec:dom-specific}, we
explore a regime detection algorithm that uses this metric.

>>>>

%  \item background: CephFS implementation
To explore software-defined cache management, we use
the data management language and policy engine presented
in~\cite{sevilla:sc15-mantle}. The prototype in that paper was called Mantle
and it was originally built on the CephFS so administrators could control file
system metadata data management policies. We now refer to Mantle as a policy
engine that supports our data management language.  The basic premise is that
data management policies can be expressed with a simple API consisting of
``when", ``where", and ``how much". ``when" controls how aggressive or
conservative the decisions are; ``where" lets the policy control how
distributed or concentrated the data should be; and ``how much" controls the
amount of data that should be sent. There is also a ``load" policy that lets
administrators specify how to collapse many metrics into a single load metric
({\it e.g.}, \(2\times\texttt{cpu} + 3\times\texttt{memory usage}\)).

The succinctness of the API lets users inject multiple, possibly dynamic, policies. In
this work we focus on a single node, so the ``where" policy is not used. When
we move ParSplice to a distributed key-value store back-end, the ``where" policy
can be used determine which key-value pairs should be moved to which node.

@@

% env of metrics
Using Mantle cluster metrics, we expose cache size, CPU utilization, and memory
pressure of the worker node to the cache management policies. In
Section~\S\ref{sec:parsplice-keyspace-analysis} we only end up using the cache
size although the other metrics proved to be valuable debugging tools. Using
Mantle time series metrics, we expose accesses to the cache as a list of
\texttt{timestamp, keyid} pairs. In Section~\S\ref{sec:dom-specific}, we
explore a key access pattern detection algorithm that uses this metric.

### 2017-11-15 paper/sections/4_dynamic-policies.tex
paper: addresses #20 and #21
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>

% why is the performance lower for smaller caches?
Although the keyspace grows to 150K, a 100K key cache achieves 99\% of the
performance. Decreasing the cache degrades performance and predictability.
Despite the memory savings, our results suggest that dynamic cache management
policies could save even more memory.  The middle plot in
Figure~\ref{fig:methodology-tradeoff} show that a 100K key cache is sufficient
as a static policy but the top graph in Figure~\ref{fig:motivation-regimes}
indicates that the cache size could be much smaller. That graph shows that the
beginning of the run is characterized by many reads to a small set of keys and
the end sees much lower reads per second to a larger keyspace. Specifically, it
shows only about 100 keys as active in the latter half of the run.

After analyzing traces, we see that the 100 key cache is insufficient because
the persistent database cannot service the read-write traffic.  According to
Figure~\ref{fig:motivation-regimes}, the read requests arrive at 750 reads per
second in addition to the writes that land in each tier (about 300 puts/second,
some redundant). This traffic triggers a LevelDB compaction and reads block,
resulting in very slow progress.  Traces verify this hypothesis and show reads
getting backed up as the read/write ratio increases. To recap, small caches
incur too much load on the persistent database  at the beginning of the run but
should suffice after the initial read flash crowd passes because
the keyspace is far less active. This suggests a two-part cache management
policy.

@@

The plot on the right side of Figure~\ref{fig:methodology-tradeoff} shows the results of using Mantle
to program a dynamic cache management policy into ParSplice that switches between
two policies:

@@

The results in Figure~\ref{fig:methodology-tradeoff} are slightly deceiving for
three reason: (1) segments take longer to generate later in the run, (2) the
memory footprint is the value at the end of 2.5 hours, and (3) this policy only
works well for the 2.5 hour run.  For (1), the curving down of the simulation
vs. wall-clock time is shown in Figure~\ref{fig:methodology-trajectory}; as the
nanoparticle grows it takes longer to generate segments so by the time we reach
2 hours, over 90\% of the trajectory is already generated.  For (2), the memory
footprint is around 0.4GB until we reach 100K key switch threshold. In
Figures~\ref{fig:methodology-tradeoff} we plot the final value. For (3),
Figure~\ref{fig:methodology-trajectory} shows that the cache fills up with 100K
keys at time 7200 seconds and its size is reduced to the size listed in the
legend.  The curves stay close to ``Unlimited" for up to an hour after the
cache is reduced but eventually flatten out as the persistent database gets
overloaded. 10K and 100K follow the ``Unlimited" curve the longest and are
sufficient policies for the 2.5 hour runs but anything longer would need a
different dynamic cache management policy.

\begin{figure}[t]
  \noindent\includegraphics[height=5cm,width=0.4\textwidth]{figures/methodology-trajectory.png}\\
  \caption{The rate that the trajectory is computed decays over time (which is
  expected) but this skews the performance improvements in the ``Multi-Part Policy" graph in 
  Figure~\ref{fig:methodology-tradeoff}. Our dynamic policy works for 2.5
  hour jobs but not for 4 hour jobs.  \label{fig:methodology-trajectory}}
\end{figure}

% but the result is still valid
Despite these caveats, the result is still valid: we found a dynamic cache 
management policy that absorbs the cost of a high read throughput on a small
keyspace and reduces the memory pressure for a 2.5 hour run. Our experiments
show the effectiveness of the policy engine we integrated into
ParSplice, not that we were able to identify the best policy for all system
setups ({\it i.e.} different ParSplice parameters, number of worker tasks, and
job lengths).  To solve that problem, we need a way to identify what thresholds
we should use for different job permutations.

@@

if #fans < 2 then return false else
  WRstate(fans[#fans-1]['top']-fans[1]['bottom'])
  return true
end
\end{minted}
\caption{The ``Dynamically Sized Cache" policy from
Figure~\ref{fig:cache-management} iterates backwards over timestamp-keyID pairs
and detects when key accesses moves on to a new subset of
keys.\label{src:dyn-cache}}
\end{figure}

>>>>

% why is the performance lower for smaller caches?
Although the keyspace grows to 150K, a 100K key cache achieves 99\% of the
performance. Decreasing the cache degrades performance and predictability.
Despite the memory savings, our results suggest that a multi-policy cache management strategy
could save even more memory.  The middle plot in
Figure~\ref{fig:methodology-tradeoff} show that a 100K key cache is sufficient
as a static policy but the top graph in Figure~\ref{fig:motivation-regimes}
indicates that the cache size could be much smaller. That graph shows that the
beginning of the run is characterized by many reads to a small set of keys and
the end sees much lower reads per second to a larger keyspace. Specifically, it
shows only about 100 keys as active in the latter half of the run.

After analyzing traces, we see that the 100 key cache is insufficient because
the persistent database cannot service the read-write traffic.  According to
Figure~\ref{fig:motivation-regimes}, the read requests arrive at 750 reads per
second in addition to the writes that land in each tier (about 300 puts/second,
some redundant). This traffic triggers a LevelDB compaction and reads block,
resulting in very slow progress.  Traces verify this hypothesis and show reads
getting backed up as the read/write ratio increases. To recap, small caches
incur too much load on the persistent database  at the beginning of the run but
should suffice after the initial read flash crowd passes because
the keyspace is far less active. This suggests a multi-policy cache management
strategy.

@@

The plot on the right side of Figure~\ref{fig:methodology-tradeoff} shows the results of using Mantle
to program a multi-policy cache management strategy into ParSplice that switches between
two policies:

@@

The results in Figure~\ref{fig:methodology-tradeoff} are slightly deceiving for
three reason: (1) segments take longer to generate later in the run, (2) the
memory footprint is the value at the end of 2.5 hours, and (3) this policy only
works well for the 2.5 hour run.  For (1), the curving down of the simulation
vs. wall-clock time is shown in Figure~\ref{fig:methodology-trajectory}; as the
nanoparticle grows it takes longer to generate segments so by the time we reach
2 hours, over 90\% of the trajectory is already generated.  For (2), the memory
footprint is around 0.4GB until we reach 100K key switch threshold. In
Figures~\ref{fig:methodology-tradeoff} we plot the final value. For (3),
Figure~\ref{fig:methodology-trajectory} shows that the cache fills up with 100K
keys at time 7200 seconds and its size is reduced to the size listed in the
legend.  The curves stay close to ``Unlimited" for up to an hour after the
cache is reduced but eventually flatten out as the persistent database gets
overloaded. 10K and 100K follow the ``Unlimited" curve the longest and are
sufficient policies for the 2.5 hour runs but anything longer would need a
different multi-policy cache management strategy.

\begin{figure}[t]
  \noindent\includegraphics[height=5cm,width=0.4\textwidth]{figures/methodology-trajectory.png}\\
  \caption{The rate that the trajectory is computed decays over time (which is
  expected) but this skews the performance improvements in the ``Multi-Part Policy" graph in 
  Figure~\ref{fig:methodology-tradeoff}. Our multi-policy cache management strategy works for 2.5
  hour jobs but not for 4 hour jobs.  \label{fig:methodology-trajectory}}
\end{figure}

% but the result is still valid
Despite these caveats, the result is still valid: we found multi-policy cache 
management strategy that absorbs the cost of a high read throughput on a small
keyspace and reduces the memory pressure for a 2.5 hour run. Our experiments
show the effectiveness of the policy engine we integrated into
ParSplice, not that we were able to identify the best policy for all system
setups ({\it i.e.} different ParSplice parameters, number of worker tasks, and
job lengths).  To solve that problem, we need a way to identify what thresholds
we should use for different job permutations.

@@

if #fans < 2 then return false else
  WRstate(fans[#fans-1]['top']-fans[1]['bottom'])
  return true
end
\end{minted}
\caption{The dynamically sized cache policy, whose performance is shown in 
Figure~\ref{fig:cache-management}, iterates backwards over timestamp-keyID pairs
and detects when key accesses move on to a new subset of
keys.\label{src:dyn-cache}}
\end{figure}

### 2017-11-15 paper/sections/5_file-systems.tex
paper: provide motivation for this work
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>
paper: addresses #20 and #21
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>

\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.48\textwidth}
        \centering
        \footnotesize
        \centering
        \begin{minted}[xleftmargin=3em,linenos]{lua}
function when()
  if server[whoami]['cachesize'] > n then
    if server[whoami]['cachesize'] > 100K then
      WRstate(1)
    end
    if RDstate() == 1 then
      return true
    end
  end
  return false
end
        \end{minted}
        \caption{ParSplice cache management policy that absorbs the burstiness of the
        workload before switching to a constrained cache.  The
        performance/utilization for different  \texttt{n} is in
        Figure~\ref{fig:methodology-tradeoff-dynamic}. \label{src:lru-dyn}}
    \end{subfigure}%
    ~ 
    \begin{subfigure}[t]{0.48\textwidth}
        \centering
        \footnotesize
        \begin{minted}[xleftmargin=3em,linenos]{lua}
local function when()
  if servers[whoami]["load"] > target then
    overloaded = RDstate() + 1
    WRstate(overloaded)
    if overloaded > 2 then
      return true
    end
  end
  else then WRstate(0) end
  return false
end
        \end{minted}
	\caption{CephFS file system metadata load balancer, designed in 2004
        in~\cite{weil:sc2004-dyn-metadata}, reimplemented in Lua
        in~\cite{sevilla:sc15-mantle}. This policy has many similarities to the
        ParSplice cache management policy.\label{src:lua-cephfs}}
    \end{subfigure}
    \caption{ParSplice's cache management policy has the same components as
    CephFS's load balancing policy.}
\end{figure*}

@@

\textbf{Two-Part Policies} (Fig.~\ref{src:lru-dyn}: Line 6;
Fig.~\ref{src:lua-cephfs}: Line 5) - after determining that the node is
overloaded, these lines add an additional condition before the policy enters a
data management state.  ParSplice trims its cache once it eclipses the
``absorb" threshold while CephFS allows balancing when overloaded for more than
two iterations. The persistent state is essential for both of these
policy-switching conditions.

@@

% burstiness of creates then compile
To explore this example in more detail, consider the trace of metadata requests
for compiling code in CephFS shown in Figure~\ref{fig:compile-ops}.  That trace
shows the number of file system metadata requests serviced by the metadata
server when uncompressing (\texttt{untar}), compiling (\texttt{make}), and
deleting (\texttt{rm}) the source code for the Linux kernel. If the system
knows that the job phases would progress from many creates, to many lookups, to
many deletes, then it could size its caches accordingly. For example, the file
system could cache none of the metadata from the \texttt{untar} phase and run
regime detection during the \texttt{make} phase, resulting in the metadata
server/clients only caching metadata that is repeatedly used. For the job in
Figure~\ref{fig:compile-ops}, this would fill up the cache to only 20K inodes
(the unit of metadata for file systems) instead of 60K, resulting in almost
40MB of memory savings (since an inode is about
1KB\footnote{http://docs.ceph.com/docs/master/dev/mds\_internals/data-structures/})
without sacrificing performance.

>>>>

\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.48\textwidth}
        \centering
        \footnotesize
        \centering
        \begin{minted}[xleftmargin=3em,linenos]{lua}
function when()
  if server[whoami]['cachesize'] > n then
    if server[whoami]['cachesize'] > 100K then
      WRstate(1)
    end
    if RDstate() == 1 then
      return true
    end
  end
  return false
end
        \end{minted}
        \caption{ParSplice cache management policy that absorbs the burstiness of the
        workload before switching to a constrained cache.  The
        performance/utilization for different  \texttt{n} is in
        Figure~\ref{fig:methodology-tradeoff}. \label{src:lru-dyn}}
    \end{subfigure}%
    ~ 
    \begin{subfigure}[t]{0.48\textwidth}
        \centering
        \footnotesize
        \begin{minted}[xleftmargin=3em,linenos]{lua}
local function when()
  if servers[whoami]["load"] > target then
    overloaded = RDstate() + 1
    WRstate(overloaded)
    if overloaded > 2 then
      return true
    end
  end
  else then WRstate(0) end
  return false
end
        \end{minted}
	\caption{CephFS file system metadata load balancer, designed in 2004
        in~\cite{weil:sc2004-dyn-metadata}, reimplemented in Lua
        in~\cite{sevilla:sc15-mantle}. This policy has many similarities to the
        ParSplice cache management policy.\label{src:lua-cephfs}}
    \end{subfigure}
    \caption{ParSplice's cache management policy has the same components as
    CephFS's load balancing policy.}
\end{figure*}

@@

\textbf{Multi-Policy Strategy} (Fig.~\ref{src:lru-dyn}: Line 6;
Fig.~\ref{src:lua-cephfs}: Line 5) - after determining that the node is
overloaded, these lines add an additional condition before the policy enters a
data management state.  ParSplice trims its cache once it eclipses the
``absorb" threshold while CephFS allows balancing when overloaded for more than
two iterations. The persistent state is essential for both of these
policy-switching conditions.

@@

% burstiness of creates then compile
To explore this example in more detail, consider the trace of metadata requests
for compiling code in CephFS shown in Figure~\ref{fig:compile-ops}.  That trace
shows the number of file system metadata requests serviced by the metadata
server when uncompressing (\texttt{untar}), compiling (\texttt{make}), and
deleting (\texttt{rm}) the source code for the Linux kernel. If the system
knows that the job phases would progress from many creates, to many lookups, to
many deletes, then it could size its caches accordingly. For example, the file
system could cache none of the metadata from the \texttt{untar} phase and run
key access pattern detection during the \texttt{make} phase, resulting in the metadata
server/clients only caching metadata that is repeatedly used. For the job in
Figure~\ref{fig:compile-ops}, this would fill up the cache to only 20K inodes
(the unit of metadata for file systems) instead of 60K, resulting in almost
40MB of memory savings (since an inode is about
1KB\footnote{http://docs.ceph.com/docs/master/dev/mds\_internals/data-structures/})
without sacrificing performance.

### 2017-11-15 paper/sections/6_conclusion.tex
paper: addresses #20 and #21
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>

Data management encompasses a wide range of techniques that vary by domain and
service. Yet, the techniques require policies that shape the decision making
and finding the best policies is a difficult, multi-dimensional problem. We
observe that many of the primitives and resulting strategies have enough in
common that they can be expressed with similar semantics. We present a data
management language and policy engine, called Mantle that is general enough to
express complicated, dynamic policies for two different domains and services.
Rather than attempting to construct a single, complex load balancing policy
that works for a variety of scenarios, we instead use the Mantle framework to
enable software-defined storage systems to flexibly change policies as the
workload changes over time.  In our analysis of the ParSplice key-value
workload we have detected clear workload regimes that are sensitive to the
initial conditions and the scale and duration of the simulation. We have also
demonstrated that changing load balancing policies at runtime in response to
the current workload is an effective mechanism to providing better load
distribution.  Finally, we have demonstrated that Mantle is flexible enough to
support domain-specific knowledge, which lays the groundwork for future work in
adaptable policies and policy generation.

>>>>

Data management encompasses a wide range of techniques that vary by domain and
service. Yet, the techniques require policies that shape the decision making
and finding the best policies is a difficult, multi-dimensional problem. We
observe that many of the primitives and resulting strategies have enough in
common that they can be expressed with similar semantics. We present a data
management language and policy engine, called Mantle that is general enough to
express complicated, multi-policy strategies for two different domains and services.
Rather than attempting to construct a single, complex load balancing policy
that works for a variety of scenarios, we instead use the Mantle framework to
enable software-defined storage systems to flexibly change policies as the
workload changes over time.  In our analysis of the ParSplice key-value
workload we have detected clear workload access patterns that are sensitive to the
initial conditions and the scale and duration of the simulation. We have also
demonstrated that changing load balancing policies at runtime in response to
the current workload is an effective mechanism to providing better load
distribution.  Finally, we have demonstrated that Mantle is flexible enough to
support domain-specific knowledge, which lays the groundwork for future work in
adaptable policies and policy generation.
